Patients with Left Hemispheric Intracerebral Hemorrhage are Less Likely to Receive Neurosurgical Intervention

### 2024-06-23 manuscripts/manuscript1.qmd
chanages

---
title: Patients with Left Hemispheric Intracerebral Hemorrhage are Less Likely to Receive Neurosurgical Intervention
author: "Nikhil Patel, MD"
subtitle: A Bayesian Re-Analysis of ATACH-2 and ERICH data
authors:
  - name: Nikhil Patel
    orcid: 0000-0002-3193-8590
    affiliation: Department of Neurology, Wake Forest University School of Medicine
    roles: data analysis, writing
    corresponding: true
  - name: Matt Jaffa
    affiliation: Department of Neurology, University of Connecticut School of Medicine
  - name: Rahul Karamchandani
    affiliation: Department of Neurology, Wake Forest University School of Medicine
  - name: Andrew Asimos
    affiliation: Department of Emergency Medicine, Wake Forest University School of Medicine
  - name: Jonathan Parish
    affiliation: Carolina Neurosurgery & Spine Associates
  - name: Nicholas Allen Morris
    affiliation: Department of Neurology, University of Maryland School of Medicine
date: last-modified
abstract: |
  Hemispheric laterality blah blah blah
table-of-contents: true
toc-title: Contents
---

@@

## Introduction
Intracerebral hemorrhage (ICH) remains a challenging disease, marked by its sudden onset, morbidity, and mortality. While ischemic stroke has garnered considerable attention and therapeutic advancements, the acute management of ICH remains limited to blood pressure control, coagulopathy reversal, and more recently surgical evacuation of hemorrhage. Overwhelmingly these efforts have focused on reducing mortality and preventing additional morbidity due to secondary injury.

@@

Previous research involving patients with ischemic stroke provide evidence that those with right hemispheric injury present later to the hospital and may not be treated as frequently with thrombolytic therapy.1,2 It has also been noted that there is a hesitation among some practitioners to offer decompressive craniectomy in patients with left dominant-hemispheric ischemic infarcts3 despite growing literature that long-term outcomes are better for patients with left hemispheric injuries in terms of both function and health related quality of life (HRQoL). [@deb-chatterjiSideMattersDifferences2022]

@@

We conducted a cohort study using data from 2 prospective studies: the Antihypertensive Treatment of Acute Cerebral Hemorrhage II (ATACH-2) trial [@qureshi_intensive_2016] and the Ethnic/Racial Variations of Intracerebral Hemorrhage (ERICH)[@woo_ethnicracial_2013] study. ATACH-2 was a large, multicenter, international, open-label, blinded-endpoint randomized controlled trial enrolling 1000 patients with ICH while ERICH was a large, multicenter, prospective observational study enrolling patients in the United States admitted with ICH. Study protocols and results of both studies have been previously published and are available elsewhere.

@@

Our exposure was hemispheric laterality (left versus right). Our primary outcome was neurosurgical intervention, defined as receiving a hematoma evacuation, decompressive craniectomy, or other craniotomy.

To explore differences in other aspects of aggressive care, we defined several secondary outcomes: receipt of external ventricular drain (EVD), tracheostomy, percutaneous endoscopic gastrostomy (PEG) tube, do-not-resuscitate (DNR) order at any point in the hospitalization, withdrawal of life-sustaining therapy (WLST) at any point in the hospitalization, and early WLST (< 72 from symptom onset).

### Statistical Analysis

#### Overall Analysis Plan

@@

A total of `r sample_size` patients were included in the analysis, of which `r left_hemisphere_count` (`r percent(left_hemisphere_count/sample_size, digits = 1)`) had left hemispheric ICH. They had higher NIHSS scores, higher GCS scores, and lower rates of hypertension as compared to those with right hemispheric ICH (@tbl-1).

>>>>

---
title: Patients with Left Hemispheric Intracerebral Hemorrhage are Less Likely to Receive Neurosurgical Intervention
author: "Nikhil Patel, MD"
subtitle: A Bayesian Re-Analysis of ATACH-2 and ERICH data
authors:
  - name: Nikhil Patel
    orcid: 0000-0002-3193-8590
    affiliation: Department of Neurology, Wake Forest University School of Medicine
    roles: data analysis, writing
    corresponding: true
  - name: Matt Jaffa
    affiliation: Department of Neurology, University of Connecticut School of Medicine
  - name: Rahul Karamchandani
    affiliation: Department of Neurology, Wake Forest University School of Medicine
  - name: Andrew Asimos
    affiliation: Department of Emergency Medicine, Wake Forest University School of Medicine
  - name: Jonathan Parish
    affiliation: Carolina Neurosurgery & Spine Associates
  - name: Nicholas Allen Morris
    affiliation: Department of Neurology, University of Maryland School of Medicine
date: last-modified
abstract: |

table-of-contents: true
toc-title: Contents
---

@@

## Introduction
Intracerebral hemorrhage (ICH) remains a challenging disease, marked by its sudden onset, morbidity, and mortality. While ischemic stroke has garnered considerable attention and therapeutic advancements, the acute management of ICH remains limited to blood pressure control, coagulopathy reversal, and more recently surgical evacuation of hemorrhage [@pradillaTrialEarlyMinimally2024]. Overwhelmingly these efforts have focused on reducing mortality and preventing additional morbidity due to secondary injury.

@@

Previous research involving patients with ischemic stroke provide evidence that those with right hemispheric injury present later to the hospital and may not be treated as frequently with thrombolytic therapy. [@dileggeImpactLesionSide2005; @foerchDifferenceRecognitionRight2005; @portegiesLeftsidedStrokesAre2015;  @liStrokeLateralizationLarge2021] It has also been noted that there is a hesitation among some practitioners to offer decompressive craniectomy in patients with left dominant-hemispheric ischemic infarcts3 despite literature demonstrating that long-term functional outcomes and health-related quality of life (HRQoL) are equivalent if not better for patients with left hemispheric injuries. [@deb-chatterjiSideMattersDifferences2022; @leggeNeglectingDifference2006; @beuscherFunctionalLongTermOutcome2017; @satoHigherMortalityPatients2015]

@@

We conducted a cohort study using data from 2 prospective studies: the Antihypertensive Treatment of Acute Cerebral Hemorrhage II (ATACH-2) trial [@qureshiIntensiveBloodPressureLowering2016] and the Ethnic/Racial Variations of Intracerebral Hemorrhage (ERICH) [@wooEthnicRacialVariations2013] study. ATACH-2 was a large, multicenter, international, open-label, blinded-endpoint randomized controlled trial enrolling 1000 patients with ICH while ERICH was a large, multicenter, prospective observational study enrolling patients in the United States admitted with ICH. Study protocols and results of both studies have been previously published and are available elsewhere.

@@

Our exposure was hemispheric laterality (left versus right). Our primary outcome was neurosurgical intervention as a binary variable defined as receiving a hematoma evacuation, decompressive craniectomy, or other craniotomy.

To explore differences in other aspects of aggressive care, we defined several secondary outcomes: receipt of external ventricular drain (EVD), tracheostomy, do-not-resuscitate (DNR) order at any point in the hospitalization, withdrawal of life-sustaining therapy (WLST) at any point in the hospitalization, and early WLST (defined as < 72 from symptom onset).

### Covariates

We constructed directed acyclic graphs (DAGs) to model our causal assumptions and to appropriately select variables to minimize confounding and selection bias. Based on our specificiations, we included the following covariates in our models: age, admission Glasgow coma score (GCS), ICH location (categorized as lobar, thalamic, or basal ganglia), baseline ICH volume (measured in millilters), and presence of intraventricular hemorrhage (IVH). We present results for both unadjusted and adjusted models.

### Statistical Analysis
To estimate the effect of hemispheric laterality on the outcomes of interest, we fitted hierarchal Bayesian regression models. This report follows the Bayesian Analysis Reporting Guidelines [@kruschkeBayesianAnalysisReporting2021]. All analyses were performed in R version 4.4.1 (R Foundation for Statistical Computing, Vienna, Austria).

#### Priors

#### Posterior distribution

We report the following metrics for the effect of hemispheric laterality on the primary and secondary outcomes: 1) median of the posterior distribution; 2) posterior distribution 95% highest density interval (HDI); 3) probability of any difference (PD; the probability that the effect size is on the side of the point estimate); 4) probability of "significance" based on a region of practical equivalence (ROPE) using traditional criteria; 5) probability of the effect size is at least equal or greater than what was considered a minimally important difference (MCID), as defined by the study authors; 6) probability of the effect size is at least 1.5 times greater than the MCID (which we considered a "large" effect).

@@

A total of `r sample_size` patients were included in the analysis, of which `r left_hemisphere_count` (`r percent(left_hemisphere_count/sample_size, digits = 1)`) had left hemispheric ICH. They had higher NIHSS scores, lower GCS scores, and lower rates of hypertension as compared to those with right hemispheric ICH (@tbl-1).
